Latent Predicate Networks: unsupervised learning of probabilistic programs with context sensitive grammars

### 2014-11-07 drafts/nips_workshop/nips_workshop.tex
new abstract, include Held-Out LL figure

%%   - learning both abstract concepts and language presents an interesting chicken and egg problem
%% 
%%   - we can model this sort of learning a conceptual grammar for sentences
%% 
%%   - importantly, this grammar must be probabilistic to accomodate noisy data and represent our uncertainty about a concept's meaning
%% 
%%   - Here, we present early work on LPNs, showing them to be the solution we want
%% 
%%   - Crucially, we show how, for certain domains, learning language and abstract concepts can be done simultaneously
%% 
%%   - We also show that this approach naturally represents sentences as probabilistic conceptual programs.
%% 
%% 
%%   While a large set of abstract concepts are necessary to understand
%%   the utterances of a language, very few of those concepts are
%%   initially available to children. Instead, they must be learned,
%%   primarily {\it via} language. One approach to understanding this
%%   joint learning problem is to model sentences as the output of
%%   probabilistic programs constructed
%% 
%% 
%% 
%%   A large set of abstract concepts are necessary to understand the
%%   utterances of a language, and at the same time, many of those
%%   abstract concepts are acquired primarily {\it via} linguistic
%%   experience. One way we might approach this sort of simultaneous
%%   learning problem is to infer the distribution over sentences, where
%%   the latent structure of this distributi
%% 
  For humans, learning abstract concepts and learning language go hand
  in hand: we acquire much of our abstract knowledge through
  linguistic experience, but the same abstract concepts are often
  crucial to interpreting linguistic expressions. How can we do both
  simultaneously? One approach is to estimate the distribution from
  which sentences are drawn, and, in doing so, infer the latent
  concepts and relationships that best explain those sentences. We
  present early work on a learning framework called Latent Predicate
  Networks (LPNs) which learns concepts by inferring the parameters of
  probabilistic context-sensitive grammars over sentences.  We show
  that for a small fragment of sentences expressing relationships
  between English number words, we can use online hierarchical Bayesian
  inference to learn LPNs that can answer simple queries about
  previously unseen relationships within this domain. These
  generalizations demonstrate LPNs' promise as a tool for learning and
  representing conceptual knowledge in language.
\end{abstract}

@@

@@

@@

First, whereas context-free grammars are suitable for describing large
swathes of language syntax, the grammars of concepts are not
context-free. To address this, we use Range Concatenation Grammars, a
context-sensitive grammar formalism -- one of several developed within
linguistics -- and extend this formalism to a probabilistic model of
sentences.

@@

Given a collection of predicates, $\{A_k\}_{k=1}^{K}$, and a
distribution over clauses, $\{\vec{w}_{A_k}\}$, the learning task is
to model a set of utterances, $\{x_j\}_{j=1}^{J}$, as being generated
according to the following distribution:

@@

In words, the weights of clauses sharing head predicate $A_k$ are
drawn from a Dirichlet distribution defined by
$\vec{\alpha}_{A_k}$. Sentences $x_j$ are then drawn from the
resulting PRCG.

Bayesian inference over stochastic grammars
and stochastic logic programs has been an active area of
research in recent decades~\cite{DBLP:journals/etai/Muggleton00, cussens2001parameter, DBLP:conf/emnlp/LiangPJK07,
  goldwater2006contextual, johnson2006adaptor}.
Variational inference is a popular approach in this domain and the one
we adopt here. As explained in the next section, we implemented
inference by translating LPNs into PRISM programs and using its
built-in Variational Bayes Expectation-Maximization
algorithm~\cite{sato2008variational}.

\section{Implementation \label{sec:implementation}}

LPNs can be encoded as a restricted subclass of PRISM programs; this
is very similar to how PCFGs are encoded in
PRISM~\cite{DBLP:conf/cl/2000}. A general probabilistic logic
programming system based on PROLOG, PRISM provides built-in predicates
for probabilistic execution and Bayesian inference over logic programs
with stochastic choices. There are several restrictions placed on
PRISM programs to maintain the validity of their probabilistic
interpretation. Most importantly, derivations of a probabilistic
predicate cannot contain cycles. Because we disallow
$A_i(\epsilon,\epsilon)$ as a valid clause, every term in the body of
a clause has shorter arguments than the head, giving acyclic and
finite derivations.

@@

\begin{figure}
    \begin{subfigure}{0.45\linewidth}
      \begin{tabular}{>{\footnotesize} l >{\footnotesize} l}
        Question & $K=4$ \\ \hline
        e.g. after twenty comes \underline{\hspace{1cm}}? & twenty one \checkmark \\
        e.g. after forty five comes \underline{\hspace{1cm}}? & forty six \checkmark \\
        e.g. after forty seven comes \underline{\hspace{1cm}}? & forty eight  \checkmark \\
        e.g. after forty nine comes \underline{\hspace{1cm}}? & forty ten $\times$ \\
        e.g. after fifty nine comes \underline{\hspace{1cm}}? & fifty ten $\times$ \\
        e.g. after sixty one comes \underline{\hspace{1cm}}? & sixty two \checkmark \\
        %% after sixty three comes \underline{\hspace{1cm}}? & sixty four \checkmark \\
        %% after sixty four comes \underline{\hspace{1cm}}? & sixty five \checkmark \\
        %% after sixty five comes \underline{\hspace{1cm}}? & sixty six \checkmark \\
        %% after sixty nine comes \underline{\hspace{1cm}}? & sixty ten $\times$ \\
        %% after seventy three comes \underline{\hspace{1cm}}? & seventy four \checkmark \\
        %% after seventy nine comes \underline{\hspace{1cm}}? & seventy ten $\times$ \\
        %% after ninety five comes \underline{\hspace{1cm}}? & ninety six \checkmark \\
        %% before twenty three comes \underline{\hspace{1cm}}? & twenty two \checkmark \\
        %% before thirty comes \underline{\hspace{1cm}}? & thirty eighty $\times$ \\
        %% before thirty eight comes \underline{\hspace{1cm}}? & thirty seven \checkmark \\
        %% before forty one comes \underline{\hspace{1cm}}? & forty \checkmark \\
        %% before fifty three comes \underline{\hspace{1cm}}? & fifty two \checkmark \\
        %% before sixty eight comes \underline{\hspace{1cm}}? & sixty seven \checkmark \\
        %% before seventy two comes \underline{\hspace{1cm}}? & seventy one \checkmark \\
        %% before seventy three comes \underline{\hspace{1cm}}? & seventy two \checkmark \\
        %% before eighty five comes \underline{\hspace{1cm}}? & eighty four \checkmark \\
        %% before ninety two comes \underline{\hspace{1cm}}? & ninety one \checkmark \\
        %% before ninety three comes \underline{\hspace{1cm}}? & ninety two \checkmark \\
        %% before ninety five comes \underline{\hspace{1cm}}? &ninety four \checkmark \\
      \end{tabular}
      \caption{Viterbi completions of held-out sentences to evaluate an LPN with 4 latent predicates in the number domain.}
      \label{tab:results}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.45\linewidth}
    \rule{6cm}{2.5cm}
    \caption{PUT HELD OUT LL FIGURE HERE}
    \label{fig:heldoutLL}
  \end{subfigure}
\end{figure}

>>>>

  Probabilistic programs successfully capture many aspects of the
  complex symbolic and statistical knowledge brought to bear in human
  cognition. Children aren't born with this knowledge, however, but
  must instead learn it, often in an unsupervised fashion. Compelling
  probabilistic programming models of human cognition must therefore
  also be learned. We present early work on Latent Predicate Networks
  (LPN), a framework for unsupervised learning in a restricted space
  of context-sensitive probabilistic programs. Specifically, we use
  Online Variational Bayesian Expectation Maximization in
  PRISM, a probabilistic programming system, to perform inference over
  the set of Stochastic Logic Programs (SLP) parseable in polynomial
  time. Using a small fragment of English sentences about number, we
  show that LPNs can recover the relevant conceptual structure
  necessary to generate novel sentences in this domain with high
  probability.

\end{abstract}

@@

%% Introduction idea
%% *** Probabilistic programs have been great models of human cognition
%% **** one-shot learning
%% **** linguistic pragmatics
%% **** some other example
%% *** one thing they lack is a good way to automatically induce full programs from data
%% *** yet, children seem to do this all the time
%% *** So, this motivates our approach
%% *** But, we have to overcome some technical challenges
%% **** many prob.prog. languages are too expressive to be tractable
%% **** many models of concept learning use too restricted a set of primitives

Probabilistic programs can successfully model many aspects of human
cognition [CITIATONS]. Part of their strength as compared to other
models of cognition lies in their ability to flexibly integrate both
symbolic and statistical information within a single framework.

One area probabilistic programming systems lag behind human cognition
is in their ability to automatically induce entire

@@

{\bf include somewhere:} A general probabilistic logic
programming system based on PROLOG, PRISM provides built-in predicates
for probabilistic execution and Bayesian inference over logic programs
with stochastic choices.

@@

First, many formal systems of probabilistic programming are so
expressive as to make learning generally intractable. To address this,
we use Range Concatenation Grammar, a context-sensitive grammar
formalism originally developed in linguistics and extend this to a
probabilistic model of programs.

@@

@@

Given a collection of predicates, $\{A_k\}_{k=1}^{K}$, and a
distribution over clauses, $\{\vec{w}_{A_k}\}$, the learning task is
to model a set of outputs, $\{x_j\}_{j=1}^{J}$, as being generated
according to the above distribution. In words, the weights of clauses
sharing head predicate $A_k$ are drawn from a Dirichlet distribution
defined by $\vec{\alpha}_{A_k}$. Each data point $x_j$ is then drawn
from the resulting PRCG. We use Bayes' Rule to update  weights $\{\vec{w}_{A_k}\}$ given our data $\vec{x}$.

\section{Inference \label{sec:implementation}}

Bayesian inference over stochastic grammars and stochastic logic
programs has been an active area of research in recent
decades~\cite{DBLP:journals/etai/Muggleton00, cussens2001parameter,
  DBLP:conf/emnlp/LiangPJK07, goldwater2006contextual,
  johnson2006adaptor}.  Variational inference is a popular approach in
this domain and the one we adopt here. We implemented inference by
translating LPNs into PRISM programs and using its built-in
Variational Bayes Expectation-Maximization
algorithm~\cite{sato2008variational}, modified for for use in an
online, rather than batch, setting. We use the adaptive learning rate
of \cite{ranganath2013adaptive}.

LPNs can be encoded as a restricted subclass of PRISM programs; this
is very similar to how PCFGs are encoded in
PRISM~\cite{DBLP:conf/cl/2000}.  There are several restrictions placed
on PRISM programs to maintain the validity of their probabilistic
interpretation. Most importantly, derivations of a probabilistic
predicate cannot contain cycles. Because we disallow
$A_i(\epsilon,\epsilon)$ as a valid clause, every term in the body of
a clause has shorter arguments than the head, giving acyclic and
finite derivations.

@@

\begin{figure}
    \begin{subfigure}[b]{0.45\linewidth}
      \begin{tabular}{>{\footnotesize} l >{\footnotesize} l}
        Question & $K=4$ \\ \hline
        XXXX after twenty comes \underline{\hspace{1cm}}? & twenty one \checkmark \\
        XXXX after forty five comes \underline{\hspace{1cm}}? & forty six \checkmark \\
        XXXX after forty seven comes \underline{\hspace{1cm}}? & forty eight  \checkmark \\
        XXXX after forty nine comes \underline{\hspace{1cm}}? & forty ten $\times$ \\
        XXXX after fifty nine comes \underline{\hspace{1cm}}? & fifty ten $\times$ \\
        XXXX after sixty one comes \underline{\hspace{1cm}}? & sixty two \checkmark \\
        XXXX after sixty three comes \underline{\hspace{1cm}}? & sixty four \checkmark \\
        XXXX after sixty four comes \underline{\hspace{1cm}}? & sixty five \checkmark \\
        XXXX after sixty five comes \underline{\hspace{1cm}}? & sixty six \checkmark \\
        XXXX after sixty nine comes \underline{\hspace{1cm}}? & sixty ten $\times$ \\
        XXXX after seventy three comes \underline{\hspace{1cm}}? & seventy four \checkmark \\
        %% after seventy nine comes \underline{\hspace{1cm}}? & seventy ten $\times$ \\
        %% after ninety five comes \underline{\hspace{1cm}}? & ninety six \checkmark \\
        %% before twenty three comes \underline{\hspace{1cm}}? & twenty two \checkmark \\
        %% before thirty comes \underline{\hspace{1cm}}? & thirty eighty $\times$ \\
        %% before thirty eight comes \underline{\hspace{1cm}}? & thirty seven \checkmark \\
        %% before forty one comes \underline{\hspace{1cm}}? & forty \checkmark \\
        %% before fifty three comes \underline{\hspace{1cm}}? & fifty two \checkmark \\
        %% before sixty eight comes \underline{\hspace{1cm}}? & sixty seven \checkmark \\
        %% before seventy two comes \underline{\hspace{1cm}}? & seventy one \checkmark \\
        %% before seventy three comes \underline{\hspace{1cm}}? & seventy two \checkmark \\
        %% before eighty five comes \underline{\hspace{1cm}}? & eighty four \checkmark \\
        %% before ninety two comes \underline{\hspace{1cm}}? & ninety one \checkmark \\
        %% before ninety three comes \underline{\hspace{1cm}}? & ninety two \checkmark \\
        %% before ninety five comes \underline{\hspace{1cm}}? &ninety four \checkmark \\
      \end{tabular}
      \caption{}
      \label{tab:results}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{figures/train_number_net_0006_held_out.pdf}
    \caption{}
    \label{fig:heldoutLL}
  \end{subfigure}
  \caption{(\subref{tab:results}) Viterbi completions of held-out sentences to evaluate an LPN with 4 latent predicates in the number domain. (\subref{fig:heldoutLL}) PUT HELD OUT LL CAPTION HERE}
\end{figure}
